Synthesising Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-09-08 papers/termination/rebuttal.txt

Reviewers 1, 3 and 4 (regarding the experimental performance): ARMC is
very good at handling loops with a small number of paths
that have linear, rational termination arguments.  We note that on
harder examples ARMC performs poorly while Juggernaut excels; for example
in the 5 benchmarks with nested loops, ARMC solves none correctly
while Juggernaut solves 4 correctly and times out on the 5th.
Since ARMC is solving a PTIME problem, it is not surprising that it
is faster than our more general algorithm in the cases that are instances
of the PTIME problem.  We also note that ARMC is not sound with respect
to bitvector semantics: our experiments include 3 cases (6%) where
ARMC terminates with an incorrect verdict.  The paper did perhaps
not make it clear that we are comparing the performance of a sound
and complete algorithm (Juggernaut) with an unsound, incomplete
algorithm (ARMC).

Reviewer 3 (regarding the comparison to SyGus): our specifications are C
code, but SyGus specifications are S-Expressions.  This design decision
allows us to use explicit state model checking and GP, as well as getting
all of compiler theory for free.  This difference is non-trivial and
novel.  However the underlying problem is clearly very similar and we
appreciate it being brought to our attention so that we may cite it.

Reviewers 2, 3 and 4 (regarding the lack of novelty over invariant
generation and other second-order techniques): we contend
that the encoding of (non-)termination as a second-order fomula is
non-trivial.  For example characterising the non-termination of nested
loops is not a straightforward extension of invariant generation: it
requires relational reasoning (which is not the same class of problem
as invariant generation -- an inherently non-relational problem) and
a novel application of Skolemisation to handle nondeterminism.
Furthermore, generating a tautology for which we find a constructive
proof is novel and, we contend, a highly non-trivial insight.
Additionally we note that previous work on using second-order logic
for program analysis does not cover termination: [19] requires
a dwf predicate in the logic to model termination, i.e. it uses a
logic with a builtin halting oracle, which makes the encoding trivial;
[20] mentions termination as an afterthought.

Reviewer 4 (regarding the importance of bitvector semantics): we agree
that people really don't *think* about termination using bitvector
arguments. If they did, our algorithm would have extremely limited value
since people would already be soundly thinking about their own code.  The
cases where a bitvector program's termination differs from its
rational counterpart are where bugs are likely to appear.  Equally,
a tool which generates a rational termination argument and checks whether
it applies for bitvector semantics has limited utility: it is likely
that if a rational argument applies, the programmer has already thought
of it.

Reviewers 2, 3 and 4 (regarding the simplicity of our approach): it is
clear that reasoning about bitvector termination is difficult, as
witnessed by the fact that ours is the first algorithm to decide it
effectively.  We consider it the main strength of our work that
we can use a sequence of simple reductions to transform a previously
intractable problem into a simple one.

Reviewer 1 (miscellaneous): C- is distinct from L in that C- is the
implementation language of part of our synthesiser, and also the language
used for specifications.
It really does not leave out any other C features.  L is the language
in which ranking functions are written and is distinct from C-.
While section 5.1 may seem excessive, we feel that many in the POPL
community are unlikely to realise that bitvector termination has no
simple relationship with rational termination: there are programs that
terminate with bitvector semantics but don't with rationls *and vice
versa*.  The ITE instruction is "if-then-else"; we agree this figure
is somewhat opaque and will add some discussion.
The runtime is predominantly a factor of the Kolmogorov complexity,
but the problem is in fact NEXPTIME-complete in the Kolmogorov complexity,
so even modest increases in complexity result in large jumps in runtime. 
We very much appreciate the suggestions regarding the rest of the
presentation and intend to incorporate them into the paper.

Reviewer 2 (miscellaneous): We do not consider our contribution to be that
we have run a lot of experiments, we consider it to be that we have
presented the first algorithm to decide bitvector termination.
We omitted the SVCOMP'15 benchmarks that made use of arrays or
recursion.  We do not have arrays in our logic and we had not implemented
recursion in our frontend (although the latter is uninteresting
theoretically).

Reviewer 3 (miscellaneous): Our contribution is twofold: we first
present a novel reduction from termination to synthesis, then we present
an algorithm and tool for solving the synthesis formula.  The first part is
unarguably significant and theoretically non-trivial (see previous
comment).  The second part is a significant algorithmic contribution
in that it manages to synthesise C programs from a C specification, and
is able to combine SAT solving, explicit state model checking and
stochastic search.  No other program synthesiser does this.

### 2014-10-17 papers/termination/termination-esop/paper.tex
esop version

\documentclass[preprint]{sigplanconf}

@@

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{pifont}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{pgf}
\usepackage[noend]{algpseudocode}
\usepackage{algorithm}
\usepackage{multicol}
\usepackage{appendix}
\usepackage{caption,subcaption}
\DeclareCaptionType{copyrightbox}
\usepackage{subfig}
\usepackage{multirow}
\usepackage{framed}
\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes, calc}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

@@

\conferenceinfo{POPL '15}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{2015}

@@

@@

Our approach is to phrase the termination and non-termination problems as
second-order satisfaction.  We identify a fragment of second-order
logic that is expressive enough to capture both problems, while being
restrictive enough to admit effective solvers.  Subsequently, we propose a method for
checking the satisfiability of a formula in this fragment by
solving an isomorphic program synthesis problem
(Section~\ref{sec:synthesis}).  Our method is sound and
complete for bit-vector programs -- for any program, we will find
a proof of either its termination or non-termination.

@@

\item We used program synthesis techniques to solve the second-order satisfaction
problems generated by our encoding.  Our solver requires only a single call to a SAT solver to verify its
proof.  It is not based on enumerating lassos, nor does it require a safety prover
that can reason about transitive closure.

\item Our solver is based on a series of source-to-source
transformations.  As well as leading to a concise implementation, this
allows us to leverage existing, mature verification tools.

\item We implemented our technique and tried it on a selection of programs
handling both bit-vectors and floats.

@@

\iffalse
\section{Combinatorics of Finite Termination}
In this section, we fix a model of computation, describe its semantics and
define the syntax of a language we will work over.

@@

Due to its expressiveness, second-order logic is very difficult to reason
in, with many second-order theories becoming undecidable even when the
corresponding first-order theory is decidable.
To make solving our constraints tractable, we identify a fragment of second-order logic
with a constrained use of quantification that is expressive enough to encode
both termination and non-termination of finite-state programs, while admitting a semi-decision procedure.
We will suggestively refer to the fragment as the \emph{synthesis fragment}:

\begin{definition}[Synthesis Fragment]
A formula is in the \emph{synthesis fragment} iff it is of the form
%
 \[
  \exists \vec{P},~ \vec{x} . ~\forall~ \vec{y} .\, \sigma(\vec{P}, \vec{x}, \vec{y})
 \]
%
where $\vec{P}$ ranges over functions, while $\vec{x}$ and $\vec{y}$ range over
ground terms.  Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times
Y^m \to \mathbb{B}$ can be seen as a specification function that returns
true iff the functions $\vec{P}$ compute appropriate outputs when fed the
inputs $\vec{x}$ and $\vec{y}$.  The synthesis fragment is implicitly
defined over some background first-order theory, and $\sigma$ must be
computable in that theory.
%
\end{definition}
%
We note that by existentially quantifying over Skolem functions, formulae with arbitrary
first-order quantification can be brought into the synthesis fragment~\cite{hol-book}, so the
fragment is semantically less restrictive than it looks.
Checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis and
%Solving the synthesis problem 
amounts to finding witnesses for the unknowns $\vec{P}$ and $\vec{x}$ that meet the specification
for all $\vec{y}$. 
If a pair $(\vec{P}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P}, \vec{x}) \models \sigma$.
For the remainder of the presentation, we drop the vector notation and write $x$ for $\vec{x}$, with the understanding
that all quantified variables range over vectors.

In the rest of this section, we show that the synthesis fragment 
is expressive enough to encode both termination and non-termination. 
Following our formulation as second-order satisfaction, 
in Section~\ref{sec:synthesis} we present a solver for the synthesis fragment that models bit-accurate semantics.

@@

We will begin our discussion by showing how to encode in the synthesis fragment the
\mbox{(non-)termination} of a program consisting of a single loop with no nesting.
For the time being, a loop $L(G, T)$ is defined by its guard $G$ and body $T$
such that states $x$ satisfying the loop's guard are given by the
predicate $G(x)$.  The body of the loop is encoded as the transition
relation $T(x, x')$, meaning that state $x'$ is reachable from state $x$ via
a single iteration of the loop body.  For example, the loop in
Figure~\ref{fig:motivation.a} is encoded as:
%
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}
We will abbreviate this with the notation:
\begin{align*}
G(x) & \triangleq x > 0 \\
T(x, x') & \triangleq x' = (x - 1) \, \& \, x
\end{align*}

@@

Returning to the program from Figure~\ref{fig:motivation.a}, we
can see that the corresponding synthesis formula {\bf [UT]} is satisfiable,
as witnessed by the function $R(x) = x$.  Thus, $R(x) = x$ constitutes a
proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
terminating.

@@

Since our program's state space is finite, a transition relation
induces an infinite execution iff some state is visited infinitely
often, or equivalently $ \exists x . T^+(x, x)$.
Deciding satisfiability of this formula directly would require a logic
that includes a transitive closure operator, $\bullet^+$.  Rather than
introduce such an operator, we will characterise non-termination
using the synthesis formula {\bf [ONT]} (Definition~\ref{def:ont}, Figure~\ref{fig:single_loop})
encoding the existence of an \emph{(open) recurrence set}, i.e.~a nonempty 
set of states $N$ such that for each $s \in N$ there
exists a transition to some $s' \in N$ \cite{DBLP:conf/popl/GuptaHMRX08}.
%% Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterise non-termination
%% of a transition relation $T$ by the existence of an \emph{(open) recurrence
%% set}, i.e.~a nonempty set of states $N$ such that for each $s \in N$ there
%% exists a transition to some $s' \in N$.
%% The notion of open recurrence set is encoded by formula {\bf [ONT]} (Definition~\ref{def:ont}).

@@

If this formula is satisfiable, $N$ is an open recurrence set for $L$, which proves
$L$'s non-termination. The issue with this formula is the additional level of quantifier alternation as compared to the synthesis fragment
(it is an $\exists \forall \exists$ formula).  To eliminate the innermost existential quantifier,
we introduce a Skolem function $C$ that chooses the successor $x'$, which we then existentially quantify over.
This results in formula {\bf [SNT]} (Definition~\ref{def:snt}, Figure~\ref{fig:single_loop}).

@@

When analysing a deterministic loop, we can make use of the notion of a \emph{closed recurrence set}
introduced by Chen et al.~in~\cite{DBLP:conf/tacas/ChenCFNO14}:  for each
state in the recurrence set $N$, \emph{all} of its successors must be in $N$.
The existence of a closed recurrence set is equivalent to the satisfiability
of formula {\bf [CNT]} in Definition~\ref{def:cnt}, which is already in the synthesis
fragment without needing Skolemization.

@@

%% Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
%% Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
%% \begin{align*}
%% W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
%% W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
%% S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
%% & z' + v' = x' \wedge v' \leq y' \\
%% R_2(x) & = x\\
%% R_3 & = ..
%% \end{align*}

@@

  i = i + 1;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.82\textwidth}
\begin{align*}
 \exists T_o, R_1, R_2 . \forall i, j, n, i', j', n' . \\
   i < n & \rightarrow T_o(\langle i, j, n \rangle , \langle i, 0, n \rangle) ~ \wedge \\
   j \leq i \wedge T_o(\langle i', j', n' \rangle, \langle i, j, n\rangle ) & \rightarrow R_2(i, j, n) > 0 ~ \wedge \\
     & R_2(i, j, n) > R_2(i, j+1, n) ~ \wedge \\
     & T_o(\langle i', j', n' \rangle, \langle i, j+1, n \rangle) ~ \wedge \\
   i < n \wedge S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow R_1(i, j, n) > 0 ~ \wedge \\
                                         & R_1(i, j, n) > R_1(i+1, j, n)
   \end{align*}
% \begin{align*}
%  \exists T_o, S, R_1, R_2 . \forall i, j, n, i', j', n' . \\
%    %% x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
%    %% x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
%    %% x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
%    S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' \leq i' & \rightarrow S(\langle i, j, n \rangle, \langle i', j'+1, n' \rangle) \, \wedge \\
%    & R_2(i,j,n) > 0 \, \wedge \\
%    & R_2(i,j,n) > R_2(i',j'+1,n') \, \wedge\\
%    i < n \wedge T_o(i, j, n) & \rightarrow S(\langle i, j, n \rangle, \langle i, 0, n \rangle) \, \wedge \\
%    i < n \wedge T_o(i, j, n) \wedge S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow \\
%    & T_o(i'+1, j', n') \, \wedge \\
%    & R_1(i, j, n) > 0 \, \wedge \\
%    & R_1(i, j, n) > R_1(i'+1, j', n')
% \end{align*}
\end{minipage}
\end{framed}
\caption{A program with nested loops and its termination formula\label{fig:environment-model}}
\end{figure*}

%% \begin{figure*}
%% \begin{framed}
%% \begin{minipage}{0.145\textwidth}
%% \begin{lstlisting}[mathescape=true]
%% x = 0;
%% $L_1:$
%% while(x<=y)
%%   x++;

%% $L_2:$
%% while(x>0){
%%   z = x;
%%   v = 0;

%% $L_3:$
%%   while(v<y){
%%     z--;
%%     v++;
%%   }

%%   x -= z;
%%   y -= z;
%% }
%% \end{lstlisting}
%% \end{minipage}
%% \vline
%% \begin{minipage}{0.85\textwidth}
%% \begin{align*}
%%  \exists W_1, W_2, R_2, R_3, S . \forall v, x, y, z, v', x', y', z' . \\
%%    x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
%%    x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
%%    x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
%%    S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' < y' & \rightarrow S(\langle v, x, y, z \rangle, \langle v'+1, x', y', z'-1 \rangle) \, \wedge \\
%%    & R_3(v,x,y,z) > 0 \, \wedge \\
%%    & R_3(v,x,y,z) > R_3(v'+1,x',y',z'-1) \, \wedge\\
%%    x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
%%    x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
%%    & W_2(v', x' - z', y' - z', z') \, \wedge \\
%%    & R_2(v, x, y, z) > 0 \, \wedge \\
%%    & R_2(v, x, y, z) > R_2(v', x' - z', y'-z', z')
%% \end{align*}
%% \end{minipage}
%% \end{framed}
%% \caption{A non-trivial program and its termination formula\label{fig:environment-model}}
%% \end{figure*}

@@

%%  \begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
%%   \begin{align*}
%% \label{def:multi-termination-formula}
%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) \rightarrow Inv_n(x_n) ~ \wedge \\
%%  & Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n') \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')
%%  \end{align*}
%% \end{definition}

%% \begin{definition}[Nested Loops Termination Formula {\bf [NT]}]
%% \label{def:nested-term-formula}
%%  \begin{align*}
%%   \exists R, S . \forall x, x' . & G_1(x) \wedge P_1(x,x') \rightarrow S(x',x') ~ \wedge \\
%%                                 & G_2(x') \wedge S(x,x') \wedge T(x',x'')\rightarrow S(x,x'') ~ \wedge \\
%% 				& G_1(x) \wedge P_1(x,x') \wedge \neg G_2(x') \wedge S(x',x'') \wedge 
%%                                 P_2(x'',x''') \wedge 
%%                                   R(x) > 0 \wedge R(x) > R(x''') 
%%  \end{align*} 
%% \end{definition}

%% \begin{definition}[Sequential Loops Non-Termination Formula {\bf [SNT]}]
%% \label{def:multi-termination-formula}
%%  \begin{align*}
%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge N(x_n)  ~ \wedge \\
%%                                                         & N(x_n) \rightarrow G_n(x_n) ~ \wedge 
%% 							 N(x_n) \wedge T_n(x_n, x_n') \rightarrow N(x_n') 
%%  \end{align*}
%% \end{definition}
 \end{framed}
\caption{Formula encoding conditional termination of a loop} \label{fig:conditional_termination}
\end{figure*}

@@

\iffalse
Let us assume that we have done some preprocessing of our program which has identified
loops, straight line code blocks and the control flow between these.  In particular,
the control flow analysis has determined which order these code blocks execute in,
and the nesting structure of the loops.
\fi

@@

%synthesizing environment abstractions: underapp for termination and underapp for non-termination.
% \cite{} addresses the problem of automatically under-approximating weakest preconditions.

%\noindent {\bf Nested loops.}
%% If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
%% {\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
%% single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
%% $L$'s environment, which we model with a relation $S$.  We require that $S$ over-approximates
%% $L'$ in the sense that it satisfies the following formula:
%% \begin{align*}
%%  \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
%% \end{align*}
%% Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
%% the inner loop ($L'$)'s transition relation.

%% To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
%% it suffices to summarise the inner loop with the relation:
%% \begin{align*}
%% S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
%% \end{align*}
%% Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.\\

%% \begin{figure}
%% \begin{lstlisting}
%% while (x > 0) {
%%   z = x, y = z + 1;
%%   while (z > 0) {
%%     z--, y--;
%%   }
%%   x -= y;
%% }
%% \end{lstlisting}
%% \caption{A nested loop\label{fig:nested}}
%% \end{figure}

% \noindent {\bf Over-approximating the environment for termination.}
%% While proving that loop $L$ terminates, we create constraints on the environment in the
%% form of the predicate $I$ and the over-approximating summary relation $T_o$:
%% In the presence of nested loops, 
%% for each loop $L'$ nested in the body
%% of $L$, we must generate constraints saying that $T_o$ over-approximates $L'$, i.e. it is an over-approximation of the transitive closure of $L'$'s transition relation.
%% Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
%% must add constraints saying that $I$ over-approximates the states reachable via $L''$.
%The details of how to create these constraints are standard, so
%rather than presenting the algorithm for generating the constraints, we direct the
%reader's attention to Figure~\ref{fig:environment-model}.  
%
% %\noindent {\bf Environment abstractions.}
% For a loop $L=(G,T)$, the formulae {\bf [CT]} %and {\bf [CNT]} 
% from Definition~\ref{def:ct} 
% %and Definition~\ref{def:cnt} 
% is in terms of 
% $T_o$ and $I$, which over-approximate the transition relation~$T$ and the set of states that $L$ may begin executing in, respectively. 
% The reason for using over-approximations [instead of exact relations] %when formulating the (non-)termination of an isolated loop 
% is so that we can use the same formulae regardless of the environment the loop is executing in, or 
% the complexity of its body, e.g.~if it contains nested loops.
% 
% \noindent {\bf Under-approximating the environment for non-termination.}

@@

Since $L$ either terminates or does not terminate, this formula is a tautology in the synthesis fragment.
A solution to the formula would include witnesses $P_N$ and $P_T$, which are putative proofs of non-termination
and termination respectively.  Exactly one of these will be a genuine proof, so we can check
first one and then the other, which will be exemplified by Algorithm~\ref{alg:termination} once
we have shown how to solve the formula {\bf [GT]}.

@@

\subsection{The Synthesis Specification}

>>>>

\documentclass[a4paper]{llncs}
%\documentclass[preprint]{sigplanconf}

@@

\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{pifont}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{pgf}
\usepackage[noend]{algpseudocode}
\usepackage{algorithm}
\usepackage{multicol}
\usepackage{appendix}
\usepackage{caption,subcaption}
\DeclareCaptionType{copyrightbox}
\usepackage{subfig}
\usepackage{multirow}
\usepackage{framed}
\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes, calc}

%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}[theorem]{Lemma}
%% \newtheorem{corollary}[theorem]{Corollary}
%% \newtheorem{conjecture}[theorem]{Conjecture}

%% \theoremstyle{definition}
%% \newtheorem{definition}[theorem]{Definition}

@@

%% \conferenceinfo{POPL '15}{Month d--d, 20yy, City, ST, Country} 
%% \copyrightyear{2015}

@@

\author{Cristina David\and Daniel Kroening\and Matt Lewis}
\institute{University of Oxford}

@@

We present a technique that treats uniformly linear and non-linear programs and
it is not restricted to finding linear ranking functions, but can
also compute lexicographic nonlinear ones. 
Our approach is constraint-based and relies on second-order formulations 
of termination and non-termination.
The obvious issue is that,
due to its expressiveness, second-order logic is very difficult to reason
in, with many second-order theories becoming undecidable even when the
corresponding first-order theory is decidable.
To make solving our constraints tractable, we formulate termination and non-termination inside a fragment of second-order logic with restricted quantification, 
for which we have built a solver in \cite{}.
Our method is sound and
complete for bit-vector programs -- for any program, we find
a proof of either its termination or non-termination.

% problems as
%second-order satisfaction.  We identify a fragment of second-order
%logic that is expressive enough to capture both problems, while being
%restrictive enough to admit effective solvers.  Subsequently, we propose a method for
%checking the satisfiability of a formula in this fragment by
%solving an isomorphic program synthesis problem
%(Section~\ref{sec:synthesis}).

@@

%% \item We used program synthesis techniques to solve the second-order satisfaction
%% problems generated by our encoding.  Our solver requires only a single call to a SAT solver to verify its
%% proof.  It is not based on enumerating lassos, nor does it require a safety prover
%% that can reason about transitive closure.

%% \item Our solver is based on a series of source-to-source
%% transformations.  As well as leading to a concise implementation, this
%% allows us to leverage existing, mature verification tools.

\item We implemented our technique and tried it on a selection of programs
handling both bit-vectors and floats. In our implementation we made use of 
a solver for second-order logic with restricted quantification that we have introduced in \cite{}.

@@

\section{Combinatorics of Finite Termination}
In this section, we fix a model of computation, describe its semantics and
define the syntax of a language we will work over.

@@

Due to its expressiveness, second-order logic is very difficult to reason
in, with many second-order theories becoming undecidable even when the
corresponding first-order theory is decidable.
In \cite{}, we have identified and built a solver for a fragment of second-order logic with restricted quantification, 
which we call second-order SAT (see Definition~\ref{def:2sat}).

\begin{definition}[Second-Order SAT]
\label{def:2sat}
 \[
  \exists S_1 \ldots S_m . Q_1 x_1 \ldots Q_n x_n . \sigma
 \]
 Where the $S_i$'s range over predicates,
the $Q_i$'s are either $\exists$ or $\forall$,
the $x_i$'s range over boolean values,
 and $\sigma$ is a quantifier-free propositional formula
 whose free variables are the $x_i$'s.  
Each $S_i$ has an associated arity $\mathrm{ar}(S_i)$
 and $S_i \in \mathbb{B}^{\mathrm{ar}(S_i)}$.  Note that 
$Q_1 x_1 \ldots Q_n x_n . \sigma$
 is an instance of first-order propositional SAT, i.e. QBF.
\end{definition}

%% To make solving our constraints tractable, we identify a fragment of second-order logic
%% with a constrained use of quantification that is expressive enough to encode
%% both termination and non-termination of finite-state programs, while admitting a semi-decision procedure.
%% We will suggestively refer to the fragment as the \emph{synthesis fragment}:

%% \begin{definition}[Synthesis Fragment]
%% A formula is in the \emph{synthesis fragment} iff it is of the form
%% %
%%  \[
%%   \exists \vec{P},~ \vec{x} . ~\forall~ \vec{y} .\, \sigma(\vec{P}, \vec{x}, \vec{y})
%%  \]
%% %
%% where $\vec{P}$ ranges over functions, while $\vec{x}$ and $\vec{y}$ range over
%% ground terms.  Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times
%% Y^m \to \mathbb{B}$ can be seen as a specification function that returns
%% true iff the functions $\vec{P}$ compute appropriate outputs when fed the
%% inputs $\vec{x}$ and $\vec{y}$.  The synthesis fragment is implicitly
%% defined over some background first-order theory, and $\sigma$ must be
%% computable in that theory.
%% %
%% \end{definition}
%
We note that by existentially quantifying over Skolem functions, formulae with arbitrary
first-order quantification can be brought into the synthesis fragment~\cite{hol-book}, so the
fragment is semantically less restrictive than it looks.
%% Checking satisfiability of a formula in second-order SAT corresponds to program synthesis and
%% amounts to finding witnesses for the unknowns $\vec{P}$ and $\vec{x}$ that meet the specification
%% for all $\vec{y}$. 
%% If a pair $(\vec{P}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P}, \vec{x}) \models \sigma$.
%% For the remainder of the presentation, we drop the vector notation and write $x$ for $\vec{x}$, with the understanding
%% that all quantified variables range over vectors.

In the rest of this section, we show that second-order SAT 
is expressive enough to encode both termination and non-termination. 
%% Following our formulation as second-order satisfaction, 
%% in Section~\ref{sec:synthesis} we present a solver for the synthesis fragment that models bit-accurate semantics.

@@

We will begin our discussion by showing how to encode in second-order SAT the
\mbox{(non-)termination} of a program consisting of a single loop with no nesting.
For the time being, a loop $L(G, T)$ is defined by its guard $G$ and body $T$
such that states $x$ satisfying the loop's guard are given by the
predicate $G(x)$.  The body of the loop is encoded as the transition
relation $T(x, x')$, meaning that state $x'$ is reachable from state $x$ via
a single iteration of the loop body.  For example, the loop in
Figure~\ref{fig:motivation.a} is encoded as:
%
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}
We will abbreviate this with the notation:
\begin{align*}
G(x) & \triangleq x > 0 \\
T(x, x') & \triangleq x' = (x - 1) \, \& \, x
\end{align*}

@@

Returning to the program from Figure~\ref{fig:motivation.a}, we
can see that the corresponding second-order SAT formula {\bf [UT]} is satisfiable,
as witnessed by the function $R(x) = x$.  Thus, $R(x) = x$ constitutes a
proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
terminating.

@@

Since our program's state space is finite, a transition relation
induces an infinite execution iff some state is visited infinitely
often, or equivalently $ \exists x . T^+(x, x)$.
Deciding satisfiability of this formula directly would require a logic
that includes a transitive closure operator, $\bullet^+$.  Rather than
introduce such an operator, we will characterise non-termination
using the second-order SAT formula {\bf [ONT]} (Definition~\ref{def:ont}, Figure~\ref{fig:single_loop})
encoding the existence of an \emph{(open) recurrence set}, i.e.~a nonempty 
set of states $N$ such that for each $s \in N$ there
exists a transition to some $s' \in N$ \cite{DBLP:conf/popl/GuptaHMRX08}.
%% Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterise non-termination
%% of a transition relation $T$ by the existence of an \emph{(open) recurrence
%% set}, i.e.~a nonempty set of states $N$ such that for each $s \in N$ there
%% exists a transition to some $s' \in N$.
%% The notion of open recurrence set is encoded by formula {\bf [ONT]} (Definition~\ref{def:ont}).

@@

If this formula is satisfiable, $N$ is an open recurrence set for $L$, which proves
$L$'s non-termination. The issue with this formula is the additional level of quantifier alternation as compared to second-order SAT
(it is an $\exists \forall \exists$ formula).  To eliminate the innermost existential quantifier,
we introduce a Skolem function $C$ that chooses the successor $x'$, which we then existentially quantify over.
This results in formula {\bf [SNT]} (Definition~\ref{def:snt}, Figure~\ref{fig:single_loop}).

@@

When analysing a deterministic loop, we can make use of the notion of a \emph{closed recurrence set}
introduced by Chen et al.~in~\cite{DBLP:conf/tacas/ChenCFNO14}:  for each
state in the recurrence set $N$, \emph{all} of its successors must be in $N$.
The existence of a closed recurrence set is equivalent to the satisfiability
of formula {\bf [CNT]} in Definition~\ref{def:cnt}, which is already in second-order SAT without needing Skolemization.

@@

@@

  i = i + 1;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.82\textwidth}
\begin{align*}
 \exists T_o, R_1, R_2 . \forall i, j, n, i', j', n' . \\
   i < n & \rightarrow T_o(\langle i, j, n \rangle , \langle i, 0, n \rangle) ~ \wedge \\
   j \leq i \wedge T_o(\langle i', j', n' \rangle, \langle i, j, n\rangle ) & \rightarrow R_2(i, j, n) > 0 ~ \wedge \\
     & R_2(i, j, n) > R_2(i, j+1, n) ~ \wedge \\
     & T_o(\langle i', j', n' \rangle, \langle i, j+1, n \rangle) ~ \wedge \\
   i < n \wedge S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow R_1(i, j, n) > 0 ~ \wedge \\
                                         & R_1(i, j, n) > R_1(i+1, j, n)
   \end{align*}
\end{minipage}
\end{framed}
\caption{A program with nested loops and its termination formula\label{fig:environment-model}}
\end{figure*}

@@

 \end{framed}
\caption{Formula encoding conditional termination of a loop} \label{fig:conditional_termination}
\end{figure*}

@@

Let us assume that we have done some preprocessing of our program which has identified
loops, straight line code blocks and the control flow between these.  In particular,
the control flow analysis has determined which order these code blocks execute in,
and the nesting structure of the loops.

@@

@@

Since $L$ either terminates or does not terminate, this formula is a tautology in second-order SAT.
A solution to the formula would include witnesses $P_N$ and $P_T$, which are putative proofs of non-termination
and termination respectively.  Exactly one of these will be a genuine proof, so we can check
first one and then the other, which will be exemplified by Algorithm~\ref{alg:termination} once
we have shown how to solve the formula {\bf [GT]}.

@@

%% \subsection{The Synthesis Specification}
